The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-05 paper/ms.tex
Add conclusion bullet point on coordinate system offsets
Fix italics throughout

\title{The Galactic Midplane Is Not a Plane: Implications for Dynamical
Analysis with {\em Gaia} Data and Beyond}

@@

Stellar actions, computed for stars from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to label and
distinguish orbits. In principle, actions have the attractive quality of being
invariant and thus good labels of a star's orbit. Such analyses are often used
to interpret data from the {\em Gaia} mission. However, inaccurate
measurements of the phase space position and uncertainties in the assumed
Galactic potential induce an error in the computed action. We show that, in
addition to these complications, a systematic bias in the phase space position
induces a phase-dependence in the actions, which we interpret as a systematic
error. An offset in the vertical
position of $\sim15\,\pc$ for a thin disk orbit or $\sim120\,\pc$ for
a thick disk orbit is sufficient to induce a $25\,\%$ systematic error in the
vertical action $J_z$ computed under the assumption of an axisymmetric
potential. The induced error distribution is non-Gaussian and bimodal, with
neither of the modes peaking on the null value. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE project. From observations that the mean vertical velocity of the Milky Way disk varies by $\sim5\,\kms$ with radius, we estimate that the true midplane variations are of order $\sim60\textup{--}90\,\pc$. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
are likely to include a systematic vertical offset depending on the volume considered. Variation in the local
standard of rest induces similar issues. The variation of the midplane must be
taken into account when performing dynamical modeling across the large
regions of the disk accessible to {\em Gaia} and future missions.

@@

\section{Introduction} \label{sec:intro}
Our understanding of the Milky Way is currently undergoing a revolution as a
result of {\em Gaia} Data Release 2 (DR2).  Recent major discoveries include
the remnants of a major merger \citep{2018ApJ...860L..11K,
2018Natur.563...85H, 2018arXiv180704290L, 2019MNRAS.482.3426M}, a phase-space
``spiral'' in the solar neighborhood \citep{2018Natur.561..360A} possibly
indicating either local substructure infall \citep{2018MNRAS.481.1501B,
2018arXiv180800451L} or bar buckling \citep{2018arXiv181109205K}, and a gap
suggestive of perturbation by a dark matter subtructure in the tidal stream
GD1 \citep{2018ApJ...863L..20P, 2018arXiv181103631B}. These discoveries all
indicate that the Milky Way's stellar distribution, which demonstrably departs
significantly from axisymmetry, is undergoing phase mixing and dynamical
interactions across a range of spatial and temporal scales.

@@

With the advent of 6D phase-space measurements over a relatively large
($\gtrsim 2$ kpc) volume from the {\em Gaia} satellite, the study of stars'
actions has gained new popularity. One reason is dimensionality reduction---an
individual stellar orbit is concisely described by three actions, as opposed
to six phase space coordinates. Second, under the assumption of a phase-mixed
system, the dynamical properties of a population of stars should be uniquely a
function of their actions and independent of the conjugate angles, allowing
investigation of the relationship between {\em orbital} properties of stars
and other intrinsic, and at least partially invariant, properties such as age
or metallicity \citep{2018ApJ...867...31B, 2018arXiv180803278T,
2018MNRAS.481.4093S, 2019arXiv190304030G, 2019arXiv190309320D,
2019MNRAS.486.1167B}. Actions also provide a convenient basis for constructing
models of the stellar distribution function
\citep[e.g.][]{1915MNRAS..76...70J, 1985ApJ...295..388V, 2017ApJ...839...61T},
or for associating stars with similar dynamical properties (e.g. to determine
membership in moving groups).

If the system being considered departs from axisymmetry in a significant
and/or non-adiabatic way, the actions computed using an axisymmetric
approximation to the true potential can exhibit cyclic dependence on the
orbital phase (or time at which the star's position and velocity are
observed), large-scale migration, or diffusion from their initial values. In
the Milky Way, stellar actions are expected to diffuse on short time scales
due to scattering with gas clouds and to evolve on longer time scales in the
case of orbits near resonances with spiral arms, bar(s), and other large scale
perturbations \citep{2014RvMP...86....1S}. For this reason, we and other
authors have used actions to study stellar scattering in the Milky Way disk
using the improved astrometry of {\em Gaia} DR2 and various age catalogues
\citep{2018ApJ...867...31B, 2018arXiv180803278T}. Actions have also been used
to study different models of spiral structure in the Milky Way
\citep{2019MNRAS.tmp..155S}. Characteristics of the distribution of stars in
the extended solar neighborhood in action space are discussed in
\citet{2019MNRAS.484.3291T}.

The transformation to action space is nonlinear in both the phase-space
position and the potential, and requires relatively strong assumptions about
the system's symmetry. However, the Galactic potential is not strictly
axisymmetric; with the vast improvement in the quality of phase-space
measurements due to {\em Gaia}, this assumption appears increasingly
inadequate \citep[e.g.][]{2018Natur.561..360A}. Even if this assumption were
close enough for many purposes, the parameters used in axisymmetric models may
be inadequately constrained by current observations.

@@

\subsection{Quantification of the Midplane Effect} \label{ssec:quant}
We quantify how a systematic error in the Galactocentric coordinate system
induces phase-dependence in the {\em observed} actions. We will consider three
orbits in the model potential described in Section~\ref{ssec:action_comp} that
are typical of stars in the stellar thin disk, stellar thick disk, and the
stellar halo. We summarize their initial positions in phase space and the
actions computed by integrating their orbits in the correct potential in
Table~\ref{tab:orbits}. Each orbit, integrated without systematic coordinate
errors, is plotted in Appendix~\ref{app:orbits}. We will refer to these orbits
by their names (thin-disk, thick-disk, halo) henceforth.

@@

We also perform the same procedure but assume a $100\,\pc$ offset in the $x$
component (i.e. subtracting the vector $(100, 0, 0)\,\pc$) in the {\em lower}
panels. This is equivalent to a systematic offset in the determination of the
distance from the Sun to the Galactic center.

@@

\subsection{Solar Height}
Third, one must define the Sun's vertical distance from the Galactic midplane,
which can be determined by identifying where the stellar density and
velocities reach a maximum (effectively the median height of all disk stars).
The Solar height is usually taken to be $\sim 25\,\pc$
\citep{2001ApJ...553..184C}, with a more recent measurement from {\em Gaia}
DR2 placing it at $20.8 \pm 0.3\,\pc$ \citep{2019MNRAS.482.1417B}. Another
strategy is to use the cold gas or H~\textsc{ii} regions in the disk to define
the Galactic midplane, leading to slightly different values (by $\sim 5\,\pc$)
for the Sun's relative height \citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em
Gaia} review of these measurements is given by \citet{2016ARAA..54..529B}.

@@

Systematic variations in $v_z$ were first noted as asymmetries in the local
velocity distribution towards the North and South Galactic Caps from the
radial velocity surveys of the Sloan Digital Sky Surveys
\citep{2012ApJ...750L..41W} and RAdial Velocity Experiment
\citep{2013MNRAS.436..101W}. Subsequently, \citet{2013ApJ...777L...5C} pointed
out suggestions of an oscillation in average vertical velocities of order
$5\,\kms$  on $\sim\,\kpc$ scales looking toward the Galactic anticenter. Work
by the {\em Gaia} collaboration confirmed these preliminary results on the
velocity and spatial scales of these oscillation with clear spatial maps made
using DR2 data of median $v_z$ over a significant Galactic volume
\citep{2018A&A...616A..11G, 2019arXiv190209569F}. We also see in the FIRE
simulation that the vertical velocity variation as a function of azimuth is
$\sim5\text{--}10\,\kms$ (Figure~\ref{fig:lsr_variations}), consistent with
these observations.

@@

The assumption that the dynamical Galactic center and \sgra{} are colocated is
tested in any construction of a dynamical model where $R_0$ is a free
parameter. For example, \citet{2015ApJ...803...80K} measured $R_0$ while
modeling the dynamics of the stream Palomar 5. Many other dynamical
measurements of $R_0$ have been made (\citealt{2016ARAA..54..529B} summarize
many pre-{\em Gaia} results), but none have yet achieved a precision
comparable to that of the distance to \sgra{}.

We did not consider in this text the effect of the {\em angular position} of
the dynamical Galactic center being offset with \sgra{}.

@@

\item The distribution of systematic errors in the actions induced by a
coordinate system offset is very non-Gaussian. The distribution is bimodal
with {\em neither mode at null}. As a result, error propagation of coordinate
system offsets is complex when considering actions, and is likely to
significantly deform the action-space distribution function.

\item Dynamical modeling across large regions of the disk, over which the
midplane location varies by more than the limits discussed above, is
susceptible to this type of systematic error, since the assumption that our
local Galactic midplane is the global Galactic midplane is not true {\em a
priori}.

@@

\item Assuming a vertical velocity variation of the Milky Way of
$\sim5\,\kms$, consistent with recent results from {\em Gaia} DR2
\citep{2018A&A...616A..11G, 2019arXiv190209569F} and our results from the FIRE
simulations (Figure~\ref{fig:lsr_variations}), we estimated that the
corresponding midplane offsets are $\sim60\textup{--}90\,\pc$ by dimensional
analysis with the vertical frequencies of disk-like orbits.

@@

Our main point is that the local midplane varies between different points in
the Galaxy, and that this variation can lead to significant systematic errors
in the computation of actions under the assumption of a global axisymmetric
potential. We do not propose here what a modeler should use for the ``true''
global midplane, since it depends on the particular problem being studied.
Even without a high-precision definition of the global Galactic midplane,
current observations from {\em Gaia} should soon permit a measurement of the
real azimuthal dependence of the midplane location. For some applications,
such as those using actions as labels to group stars on similar orbits, using
such a measurement to shift stars to a consistent midplane height as a
function of azimuth before using a global axisymmetric approximation to the
potential may be sufficient, although this ignores the \emph{dynamical}
implications of shifts in the midplane height (which result from fluctuations
in the local density). However, for other applications, such as the study of
action diffusion which motivated this paper in the first place, a more
extensive perturbative approach is likely to be needed. We plan to explore the
mitigation of these effects in future work.

>>>>

\title{The Galactic Midplane Is Not a Plane: Implications for Dynamical
Analysis with \textit{Gaia} Data and Beyond}

@@

Stellar actions, computed for stars from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to label and
distinguish orbits. In principle, actions have the attractive quality of being
invariant and thus good labels of a star's orbit. Such analyses are often used
to interpret data from the \textit{Gaia} mission. However, inaccurate
measurements of the phase space position and uncertainties in the assumed
Galactic potential induce an error in the computed action. We show that, in
addition to these complications, a systematic bias in the phase space position
induces a phase-dependence in the actions, which we interpret as a systematic
error. An offset in the vertical
position of $\sim15\,\pc$ for a thin disk orbit or $\sim120\,\pc$ for
a thick disk orbit is sufficient to induce a $25\,\%$ systematic error in the
vertical action $J_z$ computed under the assumption of an axisymmetric
potential. The induced error distribution is non-Gaussian and bimodal, with
neither of the modes peaking on the null value. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE project. From observations that the mean vertical velocity of the Milky Way disk varies by $\sim5\,\kms$ with radius, we estimate that the true midplane variations are of order $\sim60\textup{--}90\,\pc$. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
are likely to include a systematic vertical offset depending on the volume considered. Variation in the local
standard of rest induces similar issues. The variation of the midplane must be
taken into account when performing dynamical modeling across the large
regions of the disk accessible to \textit{Gaia} and future missions.

@@

\section{Introduction} \label{sec:intro}
Our understanding of the Milky Way is currently undergoing a revolution as a
result of \textit{Gaia} Data Release 2 (DR2).  Recent major discoveries include
the remnants of a major merger \citep{2018ApJ...860L..11K,
2018Natur.563...85H, 2018arXiv180704290L, 2019MNRAS.482.3426M}, a phase-space
``spiral'' in the solar neighborhood \citep{2018Natur.561..360A} possibly
indicating either local substructure infall \citep{2018MNRAS.481.1501B,
2018arXiv180800451L} or bar buckling \citep{2018arXiv181109205K}, and a gap
suggestive of perturbation by a dark matter subtructure in the tidal stream
GD1 \citep{2018ApJ...863L..20P, 2018arXiv181103631B}. These discoveries all
indicate that the Milky Way's stellar distribution, which demonstrably departs
significantly from axisymmetry, is undergoing phase mixing and dynamical
interactions across a range of spatial and temporal scales.

@@

With the advent of 6D phase-space measurements over a relatively large
($\gtrsim 2$ kpc) volume from the \textit{Gaia} satellite, the study of stars'
actions has gained new popularity. One reason is dimensionality reduction---an
individual stellar orbit is concisely described by three actions, as opposed
to six phase space coordinates. Second, under the assumption of a phase-mixed
system, the dynamical properties of a population of stars should be uniquely a
function of their actions and independent of the conjugate angles, allowing
investigation of the relationship between \emph{orbital} properties of stars
and other intrinsic, and at least partially invariant, properties such as age
or metallicity \citep{2018ApJ...867...31B, 2018arXiv180803278T,
2018MNRAS.481.4093S, 2019arXiv190304030G, 2019arXiv190309320D,
2019MNRAS.486.1167B}. Actions also provide a convenient basis for constructing
models of the stellar distribution function
\citep[e.g.][]{1915MNRAS..76...70J, 1985ApJ...295..388V, 2017ApJ...839...61T},
or for associating stars with similar dynamical properties (e.g. to determine
membership in moving groups).

If the system being considered departs from axisymmetry in a significant
and/or non-adiabatic way, the actions computed using an axisymmetric
approximation to the true potential can exhibit cyclic dependence on the
orbital phase (or time at which the star's position and velocity are
observed), large-scale migration, or diffusion from their initial values. In
the Milky Way, stellar actions are expected to diffuse on short time scales
due to scattering with gas clouds and to evolve on longer time scales in the
case of orbits near resonances with spiral arms, bar(s), and other large scale
perturbations \citep{2014RvMP...86....1S}. For this reason, we and other
authors have used actions to study stellar scattering in the Milky Way disk
using the improved astrometry of \textit{Gaia} DR2 and various age catalogues
\citep{2018ApJ...867...31B, 2018arXiv180803278T}. Actions have also been used
to study different models of spiral structure in the Milky Way
\citep{2019MNRAS.tmp..155S}. Characteristics of the distribution of stars in
the extended solar neighborhood in action space are discussed in
\citet{2019MNRAS.484.3291T}.

The transformation to action space is nonlinear in both the phase-space
position and the potential, and requires relatively strong assumptions about
the system's symmetry. However, the Galactic potential is not strictly
axisymmetric; with the vast improvement in the quality of phase-space
measurements due to \textit{Gaia}, this assumption appears increasingly
inadequate \citep[e.g.][]{2018Natur.561..360A}. Even if this assumption were
close enough for many purposes, the parameters used in axisymmetric models may
be inadequately constrained by current observations.

@@

\subsection{Quantification of the Midplane Effect} \label{ssec:quant}
We quantify how a systematic error in the Galactocentric coordinate system
induces phase-dependence in the \emph{observed} actions. We will consider three
orbits in the model potential described in Section~\ref{ssec:action_comp} that
are typical of stars in the stellar thin disk, stellar thick disk, and the
stellar halo. We summarize their initial positions in phase space and the
actions computed by integrating their orbits in the correct potential in
Table~\ref{tab:orbits}. Each orbit, integrated without systematic coordinate
errors, is plotted in Appendix~\ref{app:orbits}. We will refer to these orbits
by their names (thin-disk, thick-disk, halo) henceforth.

@@

We also perform the same procedure but assume a $100\,\pc$ offset in the $x$
component (i.e. subtracting the vector $(100, 0, 0)\,\pc$) in the lower
panels. This is equivalent to a systematic offset in the determination of the
distance from the Sun to the Galactic center.

@@

\subsection{Solar Height}
Third, one must define the Sun's vertical distance from the Galactic midplane,
which can be determined by identifying where the stellar density and
velocities reach a maximum (effectively the median height of all disk stars).
The Solar height is usually taken to be $\sim 25\,\pc$
\citep{2001ApJ...553..184C}, with a more recent measurement from \textit{Gaia}
DR2 placing it at $20.8 \pm 0.3\,\pc$ \citep{2019MNRAS.482.1417B}. Another
strategy is to use the cold gas or H~\textsc{ii} regions in the disk to define
the Galactic midplane, leading to slightly different values (by $\sim 5\,\pc$)
for the Sun's relative height \citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em
Gaia} review of these measurements is given by \citet{2016ARAA..54..529B}.

@@

Systematic variations in $v_z$ were first noted as asymmetries in the local
velocity distribution towards the North and South Galactic Caps from the
radial velocity surveys of the Sloan Digital Sky Surveys
\citep{2012ApJ...750L..41W} and RAdial Velocity Experiment
\citep{2013MNRAS.436..101W}. Subsequently, \citet{2013ApJ...777L...5C} pointed
out suggestions of an oscillation in average vertical velocities of order
$5\,\kms$  on $\sim\,\kpc$ scales looking toward the Galactic anticenter. Work
by the \textit{Gaia} collaboration confirmed these preliminary results on the
velocity and spatial scales of these oscillation with clear spatial maps made
using DR2 data of median $v_z$ over a significant Galactic volume
\citep{2018A&A...616A..11G, 2019arXiv190209569F}. We also see in the FIRE
simulation that the vertical velocity variation as a function of azimuth is
$\sim5\text{--}10\,\kms$ (Figure~\ref{fig:lsr_variations}), consistent with
these observations.

@@

The assumption that the dynamical Galactic center and \sgra{} are colocated is
tested in any construction of a dynamical model where $R_0$ is a free
parameter. For example, \citet{2015ApJ...803...80K} measured $R_0$ while
modeling the dynamics of the stream Palomar 5. Many other dynamical
measurements of $R_0$ have been made (\citealt{2016ARAA..54..529B} summarize
many pre-\textit{Gaia} results), but none have yet achieved a precision
comparable to that of the distance to \sgra{}.

We did not consider in this text the effect of the \emph{angular position} of
the dynamical Galactic center being offset with \sgra{}.

@@

\item The distribution of systematic errors in the actions induced by a
coordinate system offset is very non-Gaussian. The distribution is bimodal
with \emph{neither mode at null}. As a result, error propagation of coordinate
system offsets is complex when considering actions, and is likely to
significantly deform the action-space distribution function.

\item Dynamical modeling across large regions of the disk, over which the
midplane location varies by more than the limits discussed above, is
susceptible to this type of systematic error, since the assumption that our
local Galactic midplane is the global Galactic midplane is not true a
priori.

@@

\item Assuming a vertical velocity variation of the Milky Way of
$\sim5\,\kms$, consistent with recent results from \textit{Gaia} DR2
\citep{2018A&A...616A..11G, 2019arXiv190209569F} and our results from the FIRE
simulations (Figure~\ref{fig:lsr_variations}), we estimated that the
corresponding midplane offsets are $\sim60\textup{--}90\,\pc$ by dimensional
analysis with the vertical frequencies of disk-like orbits.

\item While inaccuracies in the parameters of the currently adopted
Galactocentric coordinate system are likely to be unimportant for most
applications, we discussed the importance of testing the assumption that the
dynamical Galactic center is colocated with \sgra{}, which we mentioned how to
do.

@@

Our main point is that the local midplane varies between different points in
the Galaxy, and that this variation can lead to significant systematic errors
in the computation of actions under the assumption of a global axisymmetric
potential. We do not propose here what a modeler should use for the ``true''
global midplane, since it depends on the particular problem being studied.
Even without a high-precision definition of the global Galactic midplane,
current observations from \textit{Gaia} should soon permit a measurement of the
real azimuthal dependence of the midplane location. For some applications,
such as those using actions as labels to group stars on similar orbits, using
such a measurement to shift stars to a consistent midplane height as a
function of azimuth before using a global axisymmetric approximation to the
potential may be sufficient, although this ignores the \emph{dynamical}
implications of shifts in the midplane height (which result from fluctuations
in the local density). However, for other applications, such as the study of
action diffusion which motivated this paper in the first place, a more
extensive perturbative approach is likely to be needed. We plan to explore the
mitigation of these effects in future work.
